Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-04-14 docs/manuscript1.Rnw
changed name of project

	\noindent\textbf{Software available from}: 
	\url{https://github.com/magnusmunch/cambridge}

@@

  Variational Bayes approximations are known to underestimate posterior 
  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
  bishop_pattern_2006,wang_inadequacy_2005}. In simulation Scenario 5 in Section 
  \ref{md-sec:simulations} of the SM, we compare the variational posterior to 
  MCMC samples from the posterior with fixed hyperparameters
  estimates (after the procedure described in Section \ref{sec:empiricalbayes}
  has converged). In this simulation Scenario and other settings (not 
  shown), the variational approximation to the posterior is accurate. If 
  however, the
  user is reluctant to trust the variational posterior variances, samples
  from the posterior may be generated with the
  Gibbs sampler in SM Section \ref{sm-sec:gibbssampler}. Alternatively, 
  we provide an
  implementation of the proposed model in stan using the R package rstan 
  \cite[]{guo_rstan:_2018} at 
  \url{https://github.com/magnusmunch/cambridge}.

@@

@@

	In all analyses we use all cell lines to estimate the hyperparameters 
	presented in Section \ref{sec:results}. Mean prediction PMSE,
	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, with
  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$, and its standard error
  are estimated by 10-fold cross validation. In the NIG and bSEM models, that 
  provide full posteriors, the posterior mean 
  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used point estimate.

	\subsection{Results}\label{sec:results}
<<analysis_gdsc>>==
# estimates tables
fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

tab1 <- t(fit)[c(1, 2), c(1:3)]
rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab1) <- c("feature intercept", "$p$-value", "drug intercept")

fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

tab2 <- t(fit)[c(1, 2), c(1:3)]
rownames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab2) <- c("feature intercept", "mutation", "drug intercept")

fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

tab3 <- t(fit)[c(1, 2), c(1:3)]
rownames(tab3) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab3) <- c("feature intercept", "$p$-value", "drug intercept")

fit <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

tab4.1 <- t(fit)[c(1, 2), c(1:2)]
rownames(tab4.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab4.1) <- c("feature intercept", "pathway")
tab4.2 <- t(fit)[c(1, 2), c(3:7)]
rownames(tab4.2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab4.2) <- c("drug intercept", 
                    "experimental", "development", "targeted", "unknown")

@@

pmse1 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse1 <- apply(pmse1, 2, mean)
sdpmse1 <- apply(pmse1, 2, sd)
ppmse1 <- paste0(round(mpmse1, 3), " (", round(sdpmse1, 3), ")")

@@

pmse2 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse2 <- apply(pmse2, 2, mean)
sdpmse2 <- apply(pmse2, 2, sd)
ppmse2 <- paste0(round(mpmse2, 3), " (", round(sdpmse2, 3), ")")

@@

pmse3 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse3 <- aggregate(pmse3, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse3 <- apply(pmse3, 2, mean)
sdpmse3 <- apply(pmse3, 2, sd)
ppmse3 <- paste0(round(mpmse3, 3), " (", round(sdpmse3, 3), ")")

@@

pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=249)), FUN="mean")[, -1]
mpmse4 <- apply(pmse4, 2, mean)
sdpmse4 <- apply(pmse4, 2, sd)
ppmse4 <- paste0(round(mpmse4, 3), " (", round(sdpmse4, 3), ")")

tab5 <- rbind(c(ppmse1[1:6], NA, ppmse1[7:length(ppmse1)], rep(NA, 3)),
              ppmse2, c(ppmse3[1:6], NA, ppmse3[7:length(ppmse3)], rep(NA, 3)),
              ppmse4)[, c(1:3, 5:7)]
tabm5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
               mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
               mpmse4)[, c(1:3, 5:7)]
tab5 <- t(sapply(1:nrow(tab5), function(r) {
  id <- which(tabm5[r, ]==min(tabm5[r, ], na.rm=TRUE))
  out <- tab5[r, ]
  out[id] <- paste0("\\textbf{", out[id], "}")
  out}))
rownames(tab5) <- paste0("analysis ", 1:nrow(tab5))
colnames(tab5) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "xtune",
                    "mxtune", "bSEM")
@
  The non-zero $\bm{\alpha}$ estimates in Tables ? and ? show 
  that there is an effect of the
  external covariates. Models NIG$_{\text{f}+\text{d}}$, NIG$_{\text{f}}$, and
  bSEM estimate a negative effect for
  the external pathway covariate. This translates to a negative additive effect 
  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
  found
  in the target pathway of a drug are more predictive of IC50 than genes that
  are not in the target pathway, according to expectation. Furthermore the 
  NIG$_{\text{f}+\text{d}}$
  model estimates positive effects for the experimental and developmental drugs,
  compared to the approved drugs. This translates to smaller expected 
  prior precisions for the regression parameters of approved drugs, i.e.,
  approved drugs are easiest to predict, followed by developmental and 
  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
  model, that only includes external
  drug covariates, estimates a negative effect for the developmental drugs.
  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
  $\bm{\beta}_d$ precision for the
  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
  are assigned the largest prior $\bm{\beta}_d$ precision, according
  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
  estimates the smallest
  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
  targeted and unkown drugs. We conjecture that, due to absence of a local
  feature-specific variance component, this model is not flexible 
  enough to correctly capture the full covariance structure of the 
  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
  prior, we calculated the range of the expected prior variances 
  (up to a multiplicative error variance) as 
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ and
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ for
  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
  estimated prior 
  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
  the external covariates have a detectable effect on the prior.

@@

  The median prediction mean squared error (PMSE), calculated on the 
  test data is displayed in Table \ref{tab:analysis_gdsc_pmse1}. 
  Note that due to standardisation, the empty reference model has a PMSE of one.
  In terms of prediction 
  error, ridge outperforms the other models 
  (PMSE of ? ), while bSEM 
  (PMSE of ? ) performs worse than 
  the null model (PMSE of ? ). Second and third best
  performing models are NIG$_{\text{f}+\text{d}}^-$ and 
  NIG$_{\text{f}+\text{d}}$ (PMSEs of ? and
  ?, respectively), both outperforming the lasso
  with a PMSE of ?. The underperformance of the 
  NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
  global-local shrinkage prior as opposed to only global or local shrinkage as 
  as in models NIG$_{\text{f}}$ and NIG$_{\text{d}}$.
<<analysis_gdsc_pmse1>>=  
kableExtra::kable_styling(knitr::kable(
  tab5, align="r",
  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data (standard error). Best performing model in bold.",
  format="latex", booktabs=TRUE, escape=FALSE), 
  latex_options=c("HOLD_position"))
@

  Part of the differences in performance between
  NIG, lasso, and ridge may be explained with the different levels of 
  ``sparsity'' in the solution. Although NIG and ridge do not give exactly 
  sparse solutions (none of the $\bm{\beta}_d$ coefficients are set to zero), we
  may use the ``heavy-tailedness'' of the distribution of the estimated 
  $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
  is excess kurtosis, calculated as ?, 
  ?, and ? for the 
  NIG$_{\text{f}+\text{d}}$, 
  lasso, and ridge estimates, respectively. Evidently, lasso gives the 
  ``sparsest'' solution, followed by NIG and ridge. In some sense, this
  means that the NIG uses less of the $\bm{\beta}_d$ to predict than the ridge
  model. In addition, the penalized regression methods estimate penalty 
  parameters by cross-validation. Cross-validation directly minimises the 
  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
  the (approximate) marginal likelihood, a measure of model fit. The superior 
  PMSE of ridge regression is therefore not surprising. A caveat with penalized
  regression methods is that they do not give measures of parameter uncertainty.
  NIG, on the other hand, gives the full posterior of the parameters, either
  through a variational Bayes approximation or with the Gibbs sampler from
  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
  to the parameter uncertainties for a better interpretable model.

  Section \ref{sm-sec:gdsc} in the SM displayes the conditional predictive 
  ordinates (CPO) for the four analyses. A visual inspection of the CPOs learns
  that no extreme outliers occur.
	\section{Discussion}\label{sec:discussion}
<<time>>=
# load("results/analysis_gdsc_fit1.Rdata")
# time2.semnig <- sum(fit2.semnig$time)
time2.semnig <- 1
@
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
	efficient and scales well with the number of features and samples.
	Estimation of the NIG model that includes both external drugs and feature
	covariates (NIG$_{\text{f}+\text{d}}$ in Section \ref{sec:gdsc}) on the GDSC
	data from Section
	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
	MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.

	Simulations in Section \ref{sec:simulations} show that the method is 
	reasonably able to estimate the hyperparameters. Simulation Scenarios 1 and 2
	show
	that estimation of drug- and feature-specific hyperparameters is, in 
	principle, fairly accurate. However, when estimated jointly, biases may occur
	due to the interplay between the two sources of information. Nevertheless,
	the simulations show that inclusion of informative external covariates is 
	still beneficial to predictive performance.

	The model is put into practice on the GDSC data. 
	The NIG model is competitive with convential models like lasso in 
	terms of predictive performance, but is slightly outperformed by ridge.
	We note that estimation of the penalty parameters in the latter two models
	is geared towards predictive performance, because cross-validation is used
	to train these parameters, as opposed to empirical Bayes estimation of the
	hyperparameters in the NIG model, which aims to improve model fit. In that 
	sense, the comparison metric (PMSE) is not to the benefit of the NIG model.
	We note that direct prediction error optimisation is not feasible in our
	external covariates setting.

	The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates covariates 
	indeed substantially modifies the hyperparameters. Although, PMSE of the NIG
	model is not necessarily lower if external covariates are included, the
	resulting hyperparameter 
	estimates are still informative and may guide future research efforts.

	In addition, the comparison to bSEM, a Bayesian counterpart model that allows
	for the inclusion of one binary external feature covariate favours the 
	proposed NIG model. PMSEs are almost uniformly lower over the drugs for the 
	NIG model compared to bSEM. Consequently, the total PMSE is significantly
	lower for the NIG model.

@@

  Another possible criticism is the treatment of $\bm{\alpha}$ as fixed
  hyperparameters instead of random. A Bayesian could argue that endowment of 
  $\bm{\alpha}$ with a hyperprior results in propagation of 
  uncertainty about $\bm{\alpha}$ and as a result improved regression parameter 
  uncertainty quantification. \cite{wiel_learning_2019} show in a similar 
  setting that EB estimation of hyperparameters does not necessarily lead to
  worse uncertainty quantification as measured by frequentist coverage of
  Bayesian credible intervals, as compared to a full Bayes treatment of the
  hyperparameters.

	A possible direction of future research to attempt
	to lower PMSE of the NIG model is the introduction of a cross-validated global
	variance component in the prior of the $\bm{\beta}_d$. If we add the 
	additional constraint that the drug and feature-specific variance components
	average to one, this cross-validated variance component may be interpreted as
	the overall level of shrinkage. This overall shrinkage level is then tailored 
	to lower PMSE, while the feature- and drug-specific variance component 
	include the external information. Other directions of future research are
	applications of the NIG
	model to different data types. One possibility is to apply the model in an 
	eQTL study, 
	where gene expressions are regressed on SNPs. Several interesting external
	covariates are available, both on the genes as well as the SNPs. An example of
	an external
	covariate for the genes is genelength, where we suspect that longer genes are
	harder to predict. The distance of the SNP to the gene is an example of an
	external SNP covariate, where the expectation is that SNPs further from the 
	gene are less predictive of that gene's expression.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{https://github.com/magnusmunch/cambridge}.

	\section*{Reproducible Research}
	All results and documents may be recreated from 
	\url{https://github.com/magnusmunch/cambridge}. All timings were obtained
	on a 2016 MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.

>>>>

	\noindent\textbf{Software available from}: 
	\url{https://github.com/magnusmunch/NIG}

@@

  Variational Bayes approximations are known to underestimate posterior 
  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
  bishop_pattern_2006,wang_inadequacy_2005}. In simulation Scenario 5 in Section 
  \ref{sm-sec:simulations} of the SM, we compare the variational posterior to 
  MCMC samples from the posterior with fixed hyperparameters
  estimates (after the procedure described in Section \ref{sec:empiricalbayes}
  has converged). In this simulation Scenario and other settings (not 
  shown), the variational approximation to the posterior is accurate. If 
  however, the
  user is reluctant to trust the variational posterior variances, samples
  from the posterior may be generated with the
  Gibbs sampler in SM Section \ref{sm-sec:gibbssampler}. Alternatively, 
  we provide an
  implementation of the proposed model in stan using the R package rstan 
  \cite[]{guo_rstan:_2018} at 
  \url{https://github.com/magnusmunch/NIG}.

@@

  Finally, the results from Scenario 5 in SM Section \ref{sm-sec:simulations} 
  show that the VB approximation is quite good as compared to standard MCMC.

@@

	In all analyses we use all cell lines to estimate the hyperparameters 
	presented in Section \ref{sec:results}. Mean prediction means squared errors
	(PMSE) and its standard error are estimated by 10-fold cross validation, where
	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, with
  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$. In the NIG and bSEM 
  models, that provide full posteriors, the posterior mean 
  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used as point estimate.

	\subsection{Results}\label{sec:results}
<<analysis_gdsc>>==
# estimates tables
fit1 <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
temp <- fit1[, 1]
fit1 <- as.matrix(fit1[, -1])
rownames(fit1) <- temp

tab1 <- t(fit1)[c(1, 2), c(1:3)]
rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab1) <- c("feature intercept", "$p$-value", "drug intercept")

fit2 <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
temp <- fit2[, 1]
fit2 <- as.matrix(fit2[, -1])
rownames(fit2) <- temp

tab2 <- t(fit2)[c(1, 2, 7), c(1:3)]
rownames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$",
                    "bSEM")
colnames(tab2) <- c("feature intercept", "mutation", "drug intercept")

fit3 <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
temp <- fit3[, 1]
fit3 <- as.matrix(fit3[, -1])
rownames(fit3) <- temp

tab3 <- t(fit3)[c(1, 2), c(1:3)]
rownames(tab3) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab3) <- c("feature intercept", "$p$-value", "drug intercept")

fit4 <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
temp <- fit4[, 1]
fit4 <- as.matrix(fit4[, -1])
rownames(fit4) <- temp

tab4.1 <- t(fit4)[c(1, 2, 7), c(1:2)]
rownames(tab4.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$", "bSEM")
colnames(tab4.1) <- c("feature intercept", "pathway")
tab4.2 <- t(fit4)[c(1, 2), c(3:7)]
rownames(tab4.2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$")
colnames(tab4.2) <- c("drug intercept", 
                    "experimental", "development", "targeted", "unknown")

@@

psel1 <- res[substr(rownames(res), 1, 5)=="psel.", ]
pmse1 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
apmse1 <- aggregate(pmse1, list(rownames(pmse1)), FUN="mean")[, -1]
pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse1 <- apply(pmse1, 2, mean)
sdpmse1 <- apply(pmse1, 2, sd)
ppmse1 <- paste0(round(mpmse1, 3), " (", round(sdpmse1, 3), ")")

@@

psel2 <- res[substr(rownames(res), 1, 5)=="psel.", ]
pmse2 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse2 <- apply(pmse2, 2, mean)
sdpmse2 <- apply(pmse2, 2, sd)
ppmse2 <- paste0(round(mpmse2, 3), " (", round(sdpmse2, 3), ")")

@@

psel3 <- res[substr(rownames(res), 1, 5)=="psel.", ]
pmse3 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse3 <- aggregate(pmse3, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
mpmse3 <- apply(pmse3, 2, mean)
sdpmse3 <- apply(pmse3, 2, sd)
ppmse3 <- paste0(round(mpmse3, 3), " (", round(sdpmse3, 3), ")")

@@

psel4 <- res[substr(rownames(res), 1, 5)=="psel.", ]
pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=249)), FUN="mean")[, -1]
mpmse4 <- apply(pmse4, 2, mean)
sdpmse4 <- apply(pmse4, 2, sd)
ppmse4 <- paste0(round(mpmse4, 3), " (", round(sdpmse4, 3), ")")

tab5 <- rbind(c(ppmse1[1:6], NA, ppmse1[7:length(ppmse1)], rep(NA, 3)),
              ppmse2, c(ppmse3[1:6], NA, ppmse3[7:length(ppmse3)], rep(NA, 3)),
              ppmse4)
tabm5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
               mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
               mpmse4)
tab5.1 <- tab5[, c(1:4, 6:7)]
tabm5.1 <- tabm5[, c(1:4, 6:7)]
tab5.1 <- t(sapply(1:nrow(tab5.1), function(r) {
  id <- which(tabm5.1[r, ]==min(tabm5.1[r, ], na.rm=TRUE))
  out <- tab5.1[r, ]
  out[id] <- paste0("\\textbf{", out[id], "}")
  out}))
rownames(tab5.1) <- paste0("analysis ", 1:nrow(tab5.1))
colnames(tab5.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", "mxtune", 
                    "bSEM")
tab5.1 <- t(tab5.1)

tab5.2 <- tab5[, c(4, 9)]
tabm5.2 <- tabm5[, c(4, 9)]
tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
  id <- which(tabm5.2[r, ]==min(tabm5.2[r, ], na.rm=TRUE))
  out <- tab5.2[r, ]
  out[id] <- paste0("\\textbf{", out[id], "}")
  out}))
rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
colnames(tab5.2) <- c("lasso", "NIG$_{\\text{f}+\\text{d}}+$DSS")
tab5.2 <- t(tab5.2)

mpsel <- rbind(c(colMeans(psel1[, 1:6]), NA, colMeans(psel1[, 7:ncol(psel1)]), 
                 rep(NA, 3)), colMeans(psel2), 
               c(colMeans(psel3[, 1:6]), NA, colMeans(psel3[, 7:ncol(psel3)]), 
                 rep(NA, 3)), colMeans(psel4))

# prior variances
vars1 <- round(1/as.numeric(model.matrix(~ factor(c(1:3))) %*% 
                              fit4[c(3, 6, 7), 2]), 4)
vars2 <- round((cbind(1, c(0, 1)) %*% fit2[c(1, 2), c(2, 7)])[1, ]/
                 (cbind(1, c(0, 1)) %*% fit2[c(1, 2), c(2, 7)])[2, ], 3)

# pmse ranges
ranges1 <- apply(rbind(mpmse1[c(1:4, 6:7)], mpmse2[c(1:4, 6:7)], 
                       mpmse3[c(1:4, 6:7)], mpmse4[c(1:4, 6:7)]), 1, range) 
ranges1 <- round(ranges1[2, ] - ranges1[1, ], 4)
psels1 <- round(mpsel[, c(4, 9)], 0)
@
  The non-zero NIG$_{\text{f}+\text{d}}$ $\bm{\alpha}$ estimates in Tables 
  \ref{tab:analysis_gdsc_fit1}-\ref{tab:analysis_gdsc_fit4.2} show 
  that there is an effect of the
  external covariates. 
  \begin{itemize}
	  \item{Analysis 1 results in a positive additive effect of the mutation 
	    $p$-values on the prior $\bm{\beta}_d$ precisions
	    (Table \ref{tab:analysis_gdsc_fit1}). This translates
	    to more $\bm{\beta}_d$ shrinkage with increasing $p$-value,
	    as expected.}
	  \item{Analysis 2 shows that gene mutations are more predictive 
	    than gene expressions, as observed from the negative effect of mutation
	    dummy on prior precions (Table \ref{tab:analysis_gdsc_fit2}): 
	    mutations are shrunken less than
	    expressions.}
	  \item{Analysis 3 indicates that CCLE $p$-values are positively related
	    to prior precision (Table \ref{tab:analysis_gdsc_fit2}), i.e., 
	    higher CCLE $p$-values results in more shrinkage of $\bm{\beta}_d$, as 
	    expected.}
	  \item{Analysis 4 gives a negative effect for the pathway dummy (Table 
	    \ref{tab:analysis_gdsc_fit4.1}), indicating less shrinkage for genes
	    that are in the drugs' target pathway, as expected. According to 
	    expectation, experimental and 
	    developmental drugs prior precisions are shrunken more than
	    the reference category, approved drugs (Table 
	    \ref{tab:analysis_gdsc_fit4.2}). Somewhat surprisingly, targeted and 
	    unkown drugs are shrunken more than the reference, cytotoxic drugs (Table 
	    \ref{tab:analysis_gdsc_fit4.2}), although the differences on the
	    $\bm{\beta}_d$ variance scale are small:
	    $\hat{\E}(\tau_{d}^{2}) = \{ \Sexpr{vars1[1]}, \Sexpr{vars1[2]}, 
	    \Sexpr{vars1[3]} \}$ for cytotoxic, targeted, and unkown drugs, 
	    respectively (ignoring other variance components).}
	\end{itemize}

  In comparison to bSEM, Analysis 2 (Table \ref{tab:analysis_gdsc_fit2}) stands
  out. The ratio of expression prior precision to mutation prior precision
  (ignoring other variance components) is
  $\hat{\E}(\tau_{\text{exp}}^{-2})/\hat{\E}(\tau_{\text{mut}}^{-2}) 
  \approx \Sexpr{vars2[1]}$ for NIG$_{\text{f}+\text{d}}$ and 
  $\hat{\E}(\tau_{\text{exp}}^{-2})/\hat{\E}(\tau_{\text{mut}}^{-2}) 
  \approx \Sexpr{vars2[2]}$ for 
  bSEM. This means that the effects are reversed: bSEM penalized mutations 
  more than expressions, while NIG$_{\text{f}+\text{d}}$ penalized
  expressions more. Furtheremore, the magnitude of the effect is much larger for
  bSEM, indicating that bSEM picks up more signal in the external covariate 
  here.

@@

  The mean prediction mean squared errors (PMSE), calculated on the 
  test data, are diplayed in Table \ref{tab:analysis_gdsc_pmse1}.
  Note that due to standardisation, an empty reference model has a PMSE of one.
  In terms of PMSE, ridge outperforms the other models in all Analyses 1, 3, and
  4, while bSEM outperforms in Analysis 2. In general however, all models
  perform very similar. The ranges of mean PMSE over all methods are
  \Sexpr{ranges1[1:3]}, and \Sexpr{ranges1[4]}, for Analyses 1-4, respectively,
  indicating that the differences in predictive performance are very small.
  Furthermore, the differences between models that include external covariates
  and their respective counterparts that do not, are small, indicating that the
  external covariates are not very informative here.
<<analysis_gdsc_pmse1>>=  
kableExtra::kable_styling(knitr::kable(
  tab5.1, align="r",
  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
  format="latex", booktabs=TRUE, escape=FALSE), 
  latex_options=c("HOLD_position"))
@

  Part of the differences in performance between
  NIG and ridge may be explained with the different levels of 
  sparsity in the solution. Although NIG does not automatically select features, 
  as opposed to the lasso,
  its $\bm{\beta}_d$ prior is heavy-tailed as compared to 
  the ridge prior 
  thus facilitates features selection. Selection of features from a sparse prior
  may be achieved through the decoupling shrinkage and selection approach (DSS)
  introduced in \cite{hahn_decoupling_2015}. We have applied DSS in our 
  analyses, where we select (approximately) the same number of features as 
  lasso. Table 
  \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE to lasso. The
  Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
  with the same number of features (\Sexpr{psels1[-4, 2]}, and 
  \Sexpr{psels1[4, 2]} features in the four analyses, respectively).

<<analysis_gdsc_pmse2>>=  
kableExtra::kable_styling(knitr::kable(
  tab5.2, align="r",
  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
  format="latex", booktabs=TRUE, escape=FALSE), 
  latex_options=c("HOLD_position"))
@

  To assess model fit, Section \ref{sm-sec:gdsc} in the SM displays the 
  conditional predictive 
  ordinates (CPO) for the NIG$_{\text{f}+\text{d}}$ model for the four analyses. 
  A visual inspection of the CPOs learns that no extreme outliers occur.

	\section{Discussion}\label{sec:discussion}
<<time>>=
times <- round(read.table("results/analysis_gdsc_time1.txt")[, 1], 0)
@
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Variational Bayes
	estimation is efficient and scales well with the number of features and 
	samples.
	Estimation of the NIG model that includes both external drugs and feature
	covariates (NIG$_{\text{f}+\text{d}}$ in Section \ref{sec:gdsc}) in the
	GDSC data analyses in Section \ref{sec:gdsc} took 
	\Sexpr{times[-1]}, and \Sexpr{times[1]} seconds on a 2016 
	MacBook Pro with 2 GHz Dual-Core Intel Core i5 processor and 8 GB of memory, 
	running macOS  10.15.1.

	Simulations in Section \ref{sec:simulations} show that the method is 
	reasonably able to estimate the hyperparameters. Simulation Scenarios 1 and 2
	show
	that estimation of drug- and feature-specific hyperparameters is, in 
	principle, fairly accurate. However, when estimated jointly, biases may occur
	due to the interplay between the two sources of information. Nevertheless,
	the simulation results in Section \ref{sm-sec:simulations} of the SM show that
	inclusion of informative external covariates is 
	still beneficial to predictive performance.

	The model is put into practice on the GDSC data in Section \ref{sec:gdsc}. 
	The NIG model is competitive with convential, penalized models like lasso in 
	terms of predictive performance, but is slightly outperformed by ridge.
	The penalized regression methods estimate penalty 
  parameters by cross-validation. Cross-validation directly minimises the 
  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
  the (approximate) marginal likelihood, a measure of model fit. The superior 
  PMSE of ridge regression is therefore not surprising. 
  However, direct prediction error optimisation by cross-validation is not 
	feasible in the external covariates setting, due to the large number of 
	hyperparameters.
  A caveat with penalized
  regression methods is that they do not give measures of parameter uncertainty.
  NIG, on the other hand, gives the full posterior of the parameters, either
  through a variational Bayes approximation or with the Gibbs sampler from
  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
  to the parameter uncertainties for a better interpretable model.

	The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates
	indeed substantially modifies the hyperparameters. Although, PMSE of the NIG
	model is not necessarily lower if external covariates are included, the
	resulting hyperparameter 
	estimates are still informative and may guide future research efforts.

@@

  A possible criticism of the NIG model is the treatment of $\bm{\alpha}$ as 
  fixed hyperparameters instead of random. A Bayesian could argue that endowment
  of $\bm{\alpha}$ with a hyperprior results in propagation of 
  uncertainty about $\bm{\alpha}$ and as a result improved regression parameter 
  uncertainty quantification. \cite{wiel_learning_2019} show in a similar 
  setting that EB estimation of hyperparameters does not necessarily lead to
  worse uncertainty quantification as measured by frequentist coverage of
  Bayesian credible intervals, as compared to a full Bayes treatment of the
  hyperparameters.

	Possible directions of future research are applications of the NIG model to 
	different data types. Suitable applications are eQTL studies, 
	in which gene expressions are regressed on SNPs. Several interesting external
	covariates are available, both on the genes as well as the SNPs. An example of
	an external
	covariate for the genes is genelength, where we suspect that longer genes are
	harder to predict. The distance of the SNP to the gene is an example of an
	external SNP covariate, where the expectation is that SNPs further from the 
	gene are less predictive of that gene's expression.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{https://github.com/magnusmunch/NIG}.

	\section*{Reproducible Research}
	All results and documents may be recreated from 
	\url{https://github.com/magnusmunch/NIG}.
